---
title: "Filtering data and evaluating the filtering proccess with naturaList"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{natutaList_clean_eval_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#"
)
```

In this article we will show how naturaList package can be used to process big data related to occurrence points of species occurrences. Specifically we will show how we can filter occurrence data accordingly to a list of specialists and [compare the effect](#comp_eval) of the filtering process regarding the differences in area of distribution of species, richness and the [niche space](#env_data) [before and after the filtering process](#comp_eval).

Before read this article, it is important to see the [Introduction to naturaList](https://avrodrigues.github.io/naturaList/articles/natutaList_vignette.html).

This procedure is done mainly through the use of `clean_eval` function. So, in summary, in this article you will learn how to:

-   [Process](#data_proc) this data to in order to adequate its format to naturaList functions, specifically `clean_eval` function;
-   [check for](*get_spec) possible specialists names;
-   [Compare](*comp_eval) the spatial occurrence, richness and environmental niche of species before and after the cleaning process.

## Installation and required packages {#install}

For this article the user must read or install the following packages:

```{r install, eval=F, echo = T}
library(naturaList)
library(devtools)
library(raster)
library(maptools)
library(rnaturalearth)
library(sf)
library(devtools)
library(rgbif)
```

## Reading data {#read_data}

We will use occurrence data of the genus *Cyathea*, that comprises a group of fern species. We used only a genus in order to simplify the cleaning process, however this procedure can be made with data bases containing a much more species.

To read data from these species we will load the data frame containing the occurrences of *Cyatheae* species in Brazil. This data set can be loaded using the following code:
```{r load_data, eval=TRUE, echo=FALSE}

data("cyathea_occ")

```


This data frame will be used in the classification process with `classify_occ` function.

## Obtain and veryfying specialist´s list {#get_spec}

It might occur that some strings in the 'determined.by' column of the occurrence datasets do not correspond to a taxonomist name. Strings as such `"Unknown"` is often included in the 'determined.by' data field obtained from GBIF. It is important (but not necessary for naturaList functions work properly) that such strings be ignored by the `classify_occ()`.

To cope with this issue, `get_det_names()` function can be used to verify which strings are not taxonomists names. This function returns all unique strings in the 'determined.by' column of the dataset. Based on this list of names, the user can create a character vector with the strings to be ignored by `classify_occ()`, providing it to the `ignore.det.names` argument. For example, lets classify the occurrences of our dataset ignoring some strings in `determined.by` column.

```{r get_det_names, eval=F, echo =TRUE}

# check out if there are strings which are not taxonomists
det_names_all <- get_det_names(cyathea_occ, freq = TRUE)
ig.names <- c("-", "Anonymous", "Unknown") # some possible characters that do not corresponds to a specialist

```

## Classification process

After identifying possible names that do not correspond to a potential specialist in the group we have some alternatives to assembly our specialist´s list. Some of them are discussed in [Rodrigues et al](https://www.biorxiv.org/content/10.1101/2020.05.26.115220v1). Here we will adopt the strategy in which the list of specialist was obtained consulting specialists in the group. The list of specialist presented here contain names of specialists in ferns from south of Brazil, and can be accessed directly from the naturaList package using `data(speciaLists)`.

Now, `data_all_occ_sub`can be used in `classify_occ` function to perform the classification process for all species at the same time.

```{r classification, eval=F, echo=F}

data("speciaLists") #specialist data frame

occ.class_all <- classify_occ(occ = cyathea_occ, spec = speciaLists, ignore.det.names = ig.names, na.rm.coords = TRUE)

```

We can see how many occurrences were classified in each one of the six confidence levels

```{r table_res, eval=FALSE, echo=TRUE}

table(occ.class_all$naturaList_levels)

```

## Defining geographical and environmental space

### Downloding additional data {#env_data}

In order to evaluate the effects of cleaning occurrence data by mantaining only occurrences made by a specialists, we will evaluate the distribution area of *Cyathea* species before and after the cleaning process. To this propouse we will need some additional data of environmental variables, since the comparison can be done using both the geographical space (occurrence points only) or environmental space (niche space). So, we will download data from worldclim platform, select two layers, bio1 and bio2, that represents, respectively, temperature and precipitation. These variables will be used to calculate the environmental space using the function `define_env_space` from naturaList package.

```{r clim_data, eval=FALSE, echo=FALSE}

bioclim <- getData('worldclim', var='bio', res=10)

# Transform occurrence data in SpatialPointsDataFrame
spdf.occ.cl <- SpatialPoints(occ.class_all[, c("decimalLongitude", "decimalLatitude")])

# redefine the extent of bioclim layers based on buffer around the occurrences
c.bioclim <- crop(bioclim, buffer(spdf.occ.cl, 100000)) # 100km buffer

# select two layers
raster.temp.prec <- c.bioclim[[c("bio1", "bio12")]]
df.temp.prec <- as.data.frame(raster.temp.prec)

```

### Defining environmental space

Now that we have data of temperature and precipitation we can obtain the environmental space necessary to compare the effects of filtering in the spatial niche based on this two variables for all three species using the function `define_env_space` in the naturaList package

```{r env_space, eval=FALSE, echo=FALSE}

### Define the environmental space for analysis
env.space <- define_env_space(df.temp.prec, buffer.size = 0.05)

```

### Defining geographical space

We also need to define the geographical space for comparison before and after the cleaning process.

```{r geo_space, eval=T, echo=T}

# delimit the geographic space
# land area
land <- ne_countries(continent = 'south america')
un.land <- unionSpatialPolygons(land, land$scalerank)

# geo space based on crop
c.geo.space <- crop(un.land, c.bioclim)


#geo space based on intersect
i.geo.space <- intersect(un.land, buffer(spdf.occ.cl, 200000))


```

## Comparing before and after cleaning process {#comp_eval}

Here we will use the function `clean_eval` to evaluate the effect of filtering process in two characteristics of occurrence of species: the species distribution, calculated as being the convex hull polygon occupied by a species; and the environmental niche, calculated as being the convex hull polygon obtained from the (at least) two environmental variables presented in the grid in which a species occur.
Both spatial distribution and niche space are compared between the data before and after the cleaning process, and the difference in area for both measure are given for each species as one of the results of `clean_eval` function.

```{r compare, eval=F, echo = F}

cl.eval_all <- clean_eval(occ.cl = occ.class_all,
                              env.space = env.space,
                              geo.space = c.geo.space,
                              r = raster.temp.prec) # cleaning process

area_remained <- cl.eval_all$area # the amount of area remained after cleaning process 



```

The area is a value that range from 0 to 1, in which 0 indicate that none of the area of distribution of the species left after the cleaning process. On the other hand, values of 1 indicate that all the area occupied by a species remains the same after the cleaning process. In this example we cleaned our data to maintain only the occurrence points of species that was identified by a specialist.

If you are working with more than one species, one example of the usage of `clean_eval` results is to plot in a map the richness containing in each grid before and after the cleaning process. This can be done with the following code:

```{r clean_evaluation, eval=F, echo = F}
rich.before.clean <- rasterFromXYZ(cbind(cl.eval_all$site.coords,
                                         cl.eval_all$rich$rich.BC))
rich.after.clean <- rasterFromXYZ(cbind(cl.eval_all$site.coords,
                                         cl.eval_all$rich$rich.AC))


```

This can be ploted in two maps and compare the effects of filtering process in the richness of community

```{r plotRich, eval=T, echo=T}

```
